---
title: "Opioid Related Hospitalization"
author: "Prepared for Yari Valle"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	error = FALSE,
	message = FALSE,
	warning = FALSE
)
```

```{r libraries}
library(tidyverse)
library(janitor)
library(lubridate)
library(knitr)
```

```{r import}
# import hospital discharge data
hdd <- read_csv(file = "S:/HIPAA Compliance/Hospital Discharge Data/HDD_Updated-05042021/2018_to_2020/coconino_2018_to_2020.csv",
                     col_types = cols(.default = "c")) %>%
  clean_names()

# import list of medical facilties in coconino county 
coco_hospitals <- read_rds("../../coconino_county_medical_list.rds")

# filter to show only hospitals
coco_hospitals_list <- coco_hospitals %>%
  filter(str_detect(subtype, "HOSPITAL"),
         county == "COCONINO") %>%
  select(facid, legalname)
```


```{r tidy}
# discharge date filter 2019 & 2020
hdd_2020 <- hdd %>%
  mutate(discharge_date = ymd(discharge_date),
         discharge_year = year(discharge_date),
         discharge_month = month(discharge_date)) %>%
  filter(discharge_year == 2019 | discharge_year == 2020)

# use list of hospitals to filter to include only hospitals in coconino county 
# show only hdd where az_fac_id matches facid 
hdd_coco <- inner_join(x = hdd_2020,
           y = coco_hospitals_list,
           by = c("az_fac_id" = "facid"))

# from 
# Moore , BJ, and ML Barrett. “Case Study: Exploring How Opioid-Related Diagnosis Codes Translate From ICD-9-CM to ICD-10-CM.” U.S. Agency for Healthcare Research and Quality, April 24, 2017. https://www.hcup-us.ahrq.gov/datainnovations/icd10_resources.jsp.
icd_10_cm <- c(#"F112",
               #"F119",
               "T400",
               "T401",
               "T402",
               "T403",
               "T404",
               "T406"
               ) %>%
  paste(collapse = "|")
```

The information contained in this report describes the number of opioid related hospital visits for Coconino County facilities during years 2019 and 2020. The hospital discharge data is provided by AZDHS. 


```{r table}
# filter the Dx variables to match the icd 10 codes of interest 
hdd_2020_opioid <- hdd_coco %>%
  filter(if_any(contains("diagnosis"), ~str_detect(.x, icd_10_cm))) %>%
  select(az_fac_id, legalname, contains(c("discharge_", "diag")))
  
# table of opioid related hospitalizations by facility and year 
total_opioid_hospitalizations <- hdd_2020_opioid %>%
  group_by(discharge_year) %>%
  count() %>%
  ungroup() 

# caption for visualization 
table_caption <- str_c("Opioid related hospitalizations in Coconino County during years 2019 and 2020.",
      sep = "")

# data table
total_opioid_hospitalizations %>%
  knitr::kable(col.names = c("Year", "Count"),
               caption = table_caption)
```

```{r plot}
# determine plot y axis upper limit
plot_ul <- total_opioid_hospitalizations  %>%
  summarise(max = (max(n)*1.25)) %>%
  as.numeric()

# plot of opioid hospitalizations
total_opioid_hospitalizations %>%
  ggplot(mapping = aes(x = discharge_year, y = n)) +
  geom_col() +
  ylim(0, plot_ul) +
  scale_x_continuous(breaks = seq(2019, 2020, 1),
                     labels = c("2019",
                                "2020")) +
  labs(title = "Opioid Related Hospital Visits in Coconino County (2019-2020)",
       caption = table_caption,
       x = "Year",
       y = "Count") +
  theme_light()
```


## Facilities

Facilities were cross-referenced with the current *[AZDHS Medical Facilities Licensing](https://azdhs.gov/licensing/medical-facilities/index.php#consumers-databases)* list by id and limited to those shown below. 

```{r list-hospital}
# data table of opioid related hospital visits by facility and year
hdd_2020_opioid %>%
  group_by(discharge_year, legalname) %>%
  count() %>%
  pivot_wider(names_from = discharge_year,
              values_from = n) %>%
  kable(col.names = c("Facility Name",
                      "2019",
                      "2020"),
        caption = "Opioid related hospital visits in Coconino County")
```


## Deaths and Related Opioid Hospitalization

```{r deaths}
hdd_coco %>%
  filter(if_any(contains("diagnosis"), ~str_detect(.x, icd_10_cm))) %>%
  select(az_fac_id, legalname, contains(c("discharge_", "diag"))) %>%
  filter(discharge_status == "20") %>%
  group_by(discharge_year, legalname) %>%
  count() %>%
  ungroup() %>%
  pivot_wider(names_from = discharge_year,
              values_from = n) %>%
  kable(col.names = c("Facility Name", "2019", "2020"),
        caption = "Count of Opioid Related Hospitalizations Where the Patient Expired")
  
```

## ICD-10 Codes

If any of the codes listed below were present in any of the diagnosis variables, then it was included in the count. Opioid related ICD-10 codes were described by:

- Moore , BJ, and ML Barrett. “Case Study: Exploring How Opioid-Related Diagnosis Codes Translate From ICD-9-CM to ICD-10-CM.” U.S. Agency for Healthcare Research and Quality, April 24, 2017. https://www.hcup-us.ahrq.gov/datainnovations/icd10_resources.jsp.


```{r list-icd}
# function to list codes 
list_function <- map(
  hdd_2020_opioid, .f = function(COL){
    tmp <- unique(COL)
    tmp[grepl(icd_10_cm, tmp)]
  }
)

# apply function 
reported_icd_10_codes <- unique(unlist(list_function))

# display results as data table 
tibble(reported_icd_10_codes) %>%
  arrange(reported_icd_10_codes) %>%
  kable(col.names = "List of ICD-10 codes included in count")
```

